---
title: "ch5 practice questions"
output: html_notebook
---

```{r setup}
require(tidyverse)
require(rethinking)
require(dagitty)
```

##5E2

mu <- a + Blxl + Bdxd

##5E3

mu <- a + Bsxs + Bfxf

Bs and Bf > 0

##5M1

Outcome: Prob of support for Kamala Harris
Predictor 1: Earned income
Predictor 2: Education

p1 and p2 are both correlated with support for Kamala Harris, but when included in the same model, only education predicts support. Earned income does not add any more information.

##5M2

Outcome: nestling growth
P1: nighttime lows
P2: daytime highs

positive correlation outcome and p1; negative correlation o and p2; positive correlation p1 and p2

##5M3

High divorce rate could make more ppl available for marriage, increasing marriage rate

Draw 2 alternative DAGs; are they Markov equivalent? If not, what are the implied conditional independencies? Test with multiple regression.

##5M4

```{r}
lds <- read_csv("../data/lds.csv")
data("WaffleDivorce")
d <- WaffleDivorce %>%
  left_join(lds, by = join_by(Location == state)) %>%
  mutate(perc_lds = str_remove(perc_lds, "%") %>% as.numeric()) %>%
  mutate(D = standardize(Divorce), 
         M = standardize(Marriage),
         A = standardize(MedianAgeMarriage),
         L = standardize(car::logit(perc_lds)))

str(d)

```

```{r}
mL <- quap(
  alist(
    D ~ dnorm(mu,sigma),
    mu <- a + bA* A  + bM * M + bL*L,
    a ~ dnorm(0,.2),
    bA ~ dnorm(0,.5),
    bM ~ dnorm(0,.5),
    bL ~ dnorm(0,.5),
    sigma ~ dexp(1)),
  data = d
)
precis(mL)

plot(coeftab(mL), par = c("bA","bM","bL"))

```

##5M5

g = price of gas
e = exercise
d = driving
r = restaurant visits
c = daily calorie consumption
o = obesity

o ~ dnorm(mu,sigma)
mu = a + be * e + bg * g and
mu = a + bg * g + bc * c
priors

So when running these two regressions in the same quap, bg should contain info about the predictor(e or c) not included in the model.

```{r}
plot(dag <- dagitty("dag{
               gas -> driving -> exercise -> obesity
               gas -> driving -> restaurantvisits -> calories -> obesity}"))

```


## 5H1

```{r}
plot(dag <- dagitty("dag{
               M -> A -> D}"))

impliedConditionalIndependencies(dag)

```

```{r}
m5.1 <- quap(
  alist(
    D ~ dnorm(mu,sigma),
    mu <- a + bA * A,
    a ~ dnorm(0,.2),
    bA ~ dnorm(0,.5),
    sigma ~ dexp(1)),
  data = d
)

m5.2 <- quap(
  alist(
    D ~ dnorm(mu,sigma),
    mu <- a + bM * M,
    a ~ dnorm(0,.2),
    bM ~ dnorm(0,.5),
    sigma ~ dexp(1)),
  data = d
)

m5.3 <- quap(
  alist(
    D ~ dnorm(mu,sigma),
    mu <- a + bA* A + bM * M,
    a ~ dnorm(0,.2),
    bA ~ dnorm(0,.5),
    bM ~ dnorm(0,.5),
    sigma ~ dexp(1)),
  data = d
)

plot(coeftab(m5.1,m5.2,m5.3), par = c("bA", "bM"))

```


yes, the data are consistent with the above implied conditional independencies of the dag.

##5H2

```{r}
m5.3_A <- quap(
  alist(
    ### A -> D
    D ~ dnorm(mu,sigma),
    mu <- a + bA* A,
    a ~ dnorm(0,.2),
    bA ~ dnorm(0,.5),
    sigma ~ dexp(1),
    ### M -> A
    A ~ dnorm(mu_A,sigma_A),
    mu_A <- aA + bMA* M,
    aA ~ dnorm(0,.2),
    bMA ~ dnorm(0,.5),
    sigma_A ~ dexp(1)
    ),
  data = d
)

precis(m5.3_A)

# With the path M -> A -> D assumed correct:

M_seq <- seq(from = -2, to = 2, length.out = 30)

sim_dat <- data.frame(M = M_seq)
s <- sim(m5.3_A, data = sim_dat, vars = c("A", "D"))

plot(sim_dat$M, colMeans(s$A), ylim = c(-2,2), type = "l",
     xlab = "manipulated M",ylab = "counterfactual A")
shade( apply(s$A, 2, PI), sim_dat$M)

plot(sim_dat$M, colMeans(s$D), ylim = c(-2,2), type = "l",
     xlab = "manipulated M",ylab = "counterfactual D")
shade( apply(s$D, 2, PI), sim_dat$M)

sim2_dat <- data.frame(M = (c(70,35) - mean(d$Marriage))/sd(d$Marriage))
s2 <- sim(m5.3_A, dat = sim2_dat, vars = c("A","D"))
mean(s2$D[,2] - s2$D[,1])

# in a perfectly controlled experiment where I can hold A constant while manipulating M:

sim_dat <- data.frame( M=seq(from=-2,to=2,length.out=30) , A=0 )
s <- sim( m5.3_A , data=sim_dat , vars="D" )
plot( sim_dat$M , colMeans(s) , ylim=c(-2,2) , type="l" ,
xlab="manipulated M" , ylab="counterfactual D" )
shade( apply(s,2,PI) , sim_dat$M )
mtext( "Total counterfactual effect of M on D" )

sim2_dat <- data.frame(M = (c(70,35) - mean(d$Marriage))/sd(d$Marriage),A = 0)
s2 <- sim(m5.3_A, dat = sim2_dat, vars = c("D"))
mean(s2[,2] - s2[,1])

```
##5H3

```{r}
data(milk)
d <- milk
str(d)


d$K <- standardize(d$kcal.per.g)
d$N <- standardize(d$neocortex.perc)
d$M <- standardize(log(d$mass))
dcc <- d[complete.cases(d$K,d$N,d$M),]

m5.7 <- quap(
  alist(
    ### M -> N <- K
    K ~ dnorm( mu , sigma ) ,
    mu <- a + bN*N + bM*M ,
    a ~ dnorm( 0 , 0.2 ) ,
    bN ~ dnorm( 0 , 0.5 ) ,
    bM ~ dnorm( 0 , 0.5 ) ,
    sigma ~ dexp( 1 ),
    ### M -> N
    N ~ dnorm(mu_N,sigma_N),
    mu_N <- aN + bMN* M,
    aN ~ dnorm(0,.2),
    bMN ~ dnorm(0,.5),
    sigma_N ~ dexp(1)
  ) , data=dcc )
precis(m5.7)

plot( coeftab( m5.7 ) , pars=c("bM","bN","bMN") )

M_seq <- seq(from = -2, to = 2, length.out = 30)

sim_dat <- data.frame(M = M_seq)
s <- sim(m5.7, data = sim_dat, vars = c("N", "K"))

plot(sim_dat$M, colMeans(s$K), ylim = c(-2,2), type = "l",
     xlab = "manipulated M",ylab = "counterfactual K")
shade( apply(s$K, 2, PI), sim_dat$M)

range(dcc$mass)
mean(dcc$mass)

sim2_dat <- data.frame(M = (c(10,20) - mean(dcc$mass))/sd(dcc$mass))
s2 <- sim(m5.7, dat = sim2_dat, vars = c("N","K"))
mean(s2$K[,2] - s2$K[,1])


```

##5H4

```{r}

lds <- read_csv("../data/lds.csv")
data("WaffleDivorce")
d <- WaffleDivorce %>%
  left_join(lds, by = join_by(Location == state)) %>%
  mutate(perc_lds = str_remove(perc_lds, "%") %>% as.numeric()) %>%
  mutate(D = standardize(Divorce), 
         M = standardize(Marriage),
         A = standardize(MedianAgeMarriage),
         L = standardize(car::logit(perc_lds)),
         S = if_else(d$South == 1,2,1))

plot(dag <- dagitty("dag{
                    A -> M
                    A -> D
                    S -> A
                    S -> M
                    S -> D}"))

impliedConditionalIndependencies(dag)

plot(dag2 <- dagitty("dag{
                    M -> A
                    A -> D
                    S -> A
                    S -> M
                    S -> D}"))

impliedConditionalIndependencies(dag2)

m5.1 <- quap(
  alist(
    D ~ dnorm(mu,sigma),
    mu <- a[S] + bA * A,
    a[S] ~ dnorm(0,.2),
    bA ~ dnorm(0,.2),
    sigma ~ dexp(1)),
  data = d
)

m5.2 <- quap(
  alist(
    D ~ dnorm(mu,sigma),
    mu <- a + bM * M,
    a ~ dnorm(0,.2),
    bM ~ dnorm(0,.5),
    sigma ~ dexp(1)),
  data = d
)

m5.3 <- quap(
  alist(
    D ~ dnorm(mu,sigma),
    mu <- a[S] + bA* A + bM * M,
    a[S] ~ dnorm(0,.2),
    bA ~ dnorm(0,.5),
    bM ~ dnorm(0,.5),
    sigma ~ dexp(1)),
  data = d
)

labels <- paste("a[",1:2,"]:",c("Not South","South",sep = ""))
plot(precis(m5.3,depth = 2, pars = c("bM","a")))
plot(coeftab(m5.2), pars = c("a","bM"))

plot(coeftab(m5.1,m5.2,m5.3), pars = c("bA", "bM", "a"))

```

Southernness may increase States' divorce rates through its influence on the other variables.

I think that Southern States will decrease age at marriage, increase overall marriage rate, and decrease divorce rate directly. However, the effect of decreasing age at marriage (increased divorce rates) seems to swamp the direct effect of Southernness (decreased divorce rate).

```{r}

m5.3 <- quap(
  alist(
    ### S, A, M -> D
    D ~ dnorm(mu,sigma),
    mu <- a[S] + bA* A + bM * M,
    a[S] ~ dnorm(0,.5),
    bA ~ dnorm(0,.5),
    bM ~ dnorm(0,.5),
    sigma ~ dexp(1),
    ### S -> A <- M
    A ~ dnorm(mu_A, sigma_A),
    mu_A <- a[S] + bMA * M,
    bMA ~ dnorm(0,.5),
    sigma_A ~ dexp(1),
    ### S -> M
    M ~ dnorm(mu_M, sigma_M),
    mu_M <- a[S],
    sigma_M ~ dexp(1)
    ),
  data = d
)

labels <- paste("a[",1:2,"]:",c("Not South","South",sep = ""))
plot(precis(m5.3,depth = 2, pars = c("a")))

post <- extract.samples(m5.3)
post$diff_s <- post$a[,1]-post$a[,2]

precis(post, depth = 2)

```

When considering the whole dag, perhaps there's no significant difference between south and not south? It seems that S might contain information already contained in A.

```{r}

m5.3 <- quap(
  alist(
    ### S, A, M -> D
    D ~ dnorm(mu,sigma),
    mu <- a[S] + bM * M,
    a[S] ~ dnorm(0,.5),
    #bA ~ dnorm(0,.5),
    bM ~ dnorm(0,.5),
    sigma ~ dexp(1),
    ### S -> A <- M
    # A ~ dnorm(mu_A, sigma_A),
    # mu_A <- a[S] + bMA * M,
    # bMA ~ dnorm(0,.5),
    # sigma_A ~ dexp(1),
    ### S -> M
    M ~ dnorm(mu_M, sigma_M),
    mu_M <- a[S],
    sigma_M ~ dexp(1)
    ),
  data = d
)

labels <- paste("a[",1:2,"]:",c("Not South","South",sep = ""))
plot(precis(m5.3,depth = 2, pars = c("a")))

post <- extract.samples(m5.3)
post$diff_s <- post$a[,1]-post$a[,2]

precis(post, depth = 2)

```

Yes. It looks like south and A contain very similar information. The effect of S on divorce rate disappears when A is included in the model.
