---
title: "ch5 practice questions"
output: html_notebook
---

```{r setup}
require(tidyverse)
require(rethinking)
```

##5E2

mu <- a + Blxl + Bdxd

##5E3

mu <- a + Bsxs + Bfxf

Bs and Bf > 0

##5M1

Outcome: Prob of support for Kamala Harris
Predictor 1: Earned income
Predictor 2: Education

p1 and p2 are both correlated with support for Kamala Harris, but when included in the same model, only education predicts support. Earned income does not add any more information.

##5M2

Outcome: nestling growth
P1: nighttime lows
P2: daytime highs

positive correlation outcome and p1; negative correlation o and p2; positive correlation p1 and p2

##5M3

High divorce rate could make more ppl available for marriage, increasing marriage rate

Draw 2 alternative DAGs; are they Markov equivalent? If not, what are the implied conditional independencies? Test with multiple regression.

##5M4

```{r}
lds <- read_csv("../data/lds.csv")
data("WaffleDivorce")
d <- WaffleDivorce %>%
  left_join(lds, by = join_by(Location == state)) %>%
  mutate(perc_lds = str_remove(perc_lds, "%") %>% as.numeric()) %>%
  mutate(D = standardize(Divorce), 
         M = standardize(Marriage),
         A = standardize(MedianAgeMarriage),
         L = standardize(perc_lds))

str(d)

```


